Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions: India's Experience Over Four Decades \bigskip Online Appendices

### 2022-03-13 paper/sex_selection_spacing-final.tex
Ensure final version can be found by Make

% For word counting remove comment on the line below
% \raggedright

% Background
India has experienced many positive changes over the past four decades:
The economy has grown substantially,
real wages have more than doubled,
educational attainment has increased for males and females,
and the total fertility rate has fallen to 2.2 children
\citep{Bosworth2008,Dharmalingam2014,Klasen2015,
International-Institute-for-Population-Sciences-IIPS2017}.

@@

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, past research has failed to appreciate that the use of sex-selective abortions could 
substantially increase birth spacing. 
As detailed below, after an abortion it takes 6 months or more to reach the same point in 
the next pregnancy. 
Second, other major societal changes, such as women's greater educational attainment, higher 
household income, and a low and declining female labor force participation, all 
likely influence birth spacing. 
The combination of the apparent increasing use of sex selection and societal 
changes can significantly impact birth spacing, but we know little about how much.

@@

% How the study progresses
To investigate how birth spacing has changed, I use a competing risk hazard model with 
two exit states: The birth of a girl or the birth of a boy. 
I apply the model to Hindu women's birth histories covering 1972--2016 using data from the 
four National Family and Health Surveys (NFHS).

@@

% Main results
There are three main results.

@@

% Economic conditions
The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility, although the direction is ambiguous; empirically, 
higher female wages reduce fertility, whereas higher male wages increase fertility 
\citep{Hotz1997,schultz97}.
According to economic theory, the substitution and income effects' relative strengths 
determine the effect of higher wages on fertility.
The substitute effect captures that when wages increase, time becomes more expensive, and
people, therefore, work more and spend less time on non-wage earning activities, such as
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Because women spend substantially more time on childrearing than men, the substitution
effect dominates for women while the income effect dominates for men.

@@

% Investment in children
With increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, thereby increasing the cost of having children and lowering fertility. 
With lower expected mortality, parents need fewer births to reach the desired number of
surviving children, which, in turn, allows parents to invest more in each child.

@@

% Son preference 
Stronger son preference has traditionally been associated with shorter spacing and worse
health outcomes for daughters when there no sons in the household
\citep{Whitworth2002,Jayachandran2011}.
Although son preference increases average fertility only marginally, differential stopping
behavior means that girls tend to end up in larger families, resulting in fewer resources
per child \citep{repetto72,clark00,Basu2010,Barcellos2014}.

@@

% Factors/predictions
The closest available proxies in the data for capturing the explanations linking
fertility and birth spacing are the sex composition of previous children, mothers'
education, and area of residence. 
The remainder of the section discusses how the three variables fit into the
explanations and the predicted effects.

@@

% Sanskritization
Finally, the process of "Sanskritization" implies that as lower-caste females gain access
to education, they adopt higher-caste norms such as stronger son preference and a
retraction from the formal labor market \citep{Srinivas1956}. 
The declining female labor force participation suggests that this process still 
operates \citep{Abraham2013,Chatterjee2018}.
Thus, although the expansion in female education will likely bring in groups who initially 
have different norms, we are unlikely to see substantial changes in how women with a given 
education level behave.

% Summary / hypotheses
In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated as their
household income increase the most, and because of their presumptive use of sex selection. 
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.

@@

% Requirements for method and the intuition behind each:  
%  - Account for censoring (hazard model)
%  - Work both with and without sex selection (competing risk part)
%  - Allow for changes in use of sex selection within spell (flexible specification of baseline hazard ?)
%  - Capture that the shape of the hazard function differ across groups
%    (nonproportionality); needs a better description, maybe using example
%    of shorter spacing and/or use of sex selection. Differences in parity
%    progression likelihood.

@@

% To further minimize any potential bias from assuming proportionality, estimations 
% are done separately for different levels of mothers' education and different periods.
% I discuss the choice of which variables to use for nonproportionality below.

Equation \ref{eq:hazard} is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, splitting spells into smaller intervals---here equal to three months---and 
treating them as observations, I can estimate the model using a standard multinomial 
logit model. 
% In the reorganized data, the outcome variable is 0 if the woman does not have a child 
% in a given interval (the base outcome), 1 if she gives birth to a son in that interval, 
% and 2 if she gives birth to a daughter in that interval.

@@

For birth interval lengths, I estimate a set of percentile durations. 
I first calculate for each woman when there is a given percentage chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80\% parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60\% $(100-\frac{80}{2})$ 
mark on her survival curve. 
I then add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

% The predicted sex ratio is the weighted average of individual predicted sex ratios, using 
% parity progression probabilities as weights.
% For an individual woman, the estimated percent of births that are boys at $t$, $\hat{Y_t}$, is:  
% \begin{equation}
% \hat{Y}_t 
% = 
% \frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
% { P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
% \times 100,
% \label{eq:probability_son}
% \end{equation}
% where $b$ and $g$ are the predicted probabilities of having a boy and a girl, respectively,
% conditional on the explanatory variables and not having had a child before $t$.
% Weighting the percentages boys with the likelihood of exiting the spell with a birth across all 
% $t$ gives the predicted percent of boys over the entire spell for an individual.%
% \footnote{
% Imagine $T=2$. 
% If 54 and 66 percent of births are boys and the likelihood of giving birth 20 and 40 percent, 
% then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent boys. 
% }

@@

To ensure that there are enough births for the method to work, I censor spells at 96 
months (eight years) after a woman can first give birth, equivalent to 105 months 
after the birth of the prior child.
Less than 1\% of observed births occur after the cutoff.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

% %
% \footnote{
% \input{../tables/num_missed.tex}
% }

Direct information on sex selection is not available, so I compare periods 
based on the changes in access and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--1983, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
% Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2019}.
}
Finally, although sex selection increased even after 1994, we may have passed a 
turning point in its use in the mid-2000s \citep{Das_Gupta2009,Kumar2012,Bongaarts2013,Diamond-Smith2015}.

@@

The second group of variables consists of those expected to have an approximately 
proportional effect on the hazard.
These include the mother's age when the spell begins, the household's
land ownership, and whether it belongs to a scheduled tribe or caste.
Appendix Tables \ref{tab:des_stat1} and \ref{tab:des_stat2} present the descriptive 
statistics.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
I begin with a broad outline of how parity progression probabilities and
sex ratios have changed over time and by groups.
I then use these results to separately discuss the changes in birth spacing for 
women who do not appear to use sex selection and for those who do.

@@

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_urban}
% \caption{Percentile birth intervals, sex ratios, and parity progression  
% for urban women with no education by spell, sex composition, and period}
% \label{fig:spacing_low_urban}
% \end{figure}

@@

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_highest_rural}
% \caption{Percentile birth intervals, sex ratios, and parity progression  
% for rural women with 12 or more years of education by spell, sex composition, and period}
% \label{fig:spacing_highest_rural}
% \end{figure}

@@

% [Parity progression probability]
The parity progression and the sex ratio show two broad trends.
First, 
in line with the falling total fertility rate, the likelihood of a next birth 
has decreased over time.
The likelihood of a next birth fell more rapidly, the higher the education and the 
higher the parity.
Within a given spell and period, parity progressions are lower in urban than rural areas, 
if at least one son is present, and the more educated the mother.

% [Sex ratios]
Second, sex ratios have become more male-dominated for women with no sons, indicating the spread 
of sex selection. 
The percentage of boys increased more quickly, the higher the education and the higher the parity. 
There are no clear trends for the other sex compositions. 
Within a given spell and period combination and in the absence of a son, sex ratios are higher, 
the more educated the mother, and the higher the parity. Sex ratios are also higher in urban than 
in rural areas.
Some women with one son also show an unnaturally high percentage of boys, although the 
failing fertility makes these estimates noisy.

@@

% [25th percentile birth intervals]
A remarkably high proportion of birth intervals are still very short.
For all but the most educated, 25\% or more have their second and third child 
within 24 months of the previous birth.
These intervals are substantially below the 24 months between 
\emph{pregnancies} the WHO recommends.
Even with the more substantial lengthening of birth intervals for higher parities, the 25th
percentile duration is still around 24 months for the fourth spell for women with less than
eight years of education.

% [median birth intervals]
Median birth interval lengths have also increased relatively little---only three to six months 
over the four decades---compared to around 3.5 months \emph{per decade} in other countries with 
declining fertility \citep{Rutstein2011,Casterline2016}.%
\footnote{
The NFHS reports show median durations of closed birth intervals of approximately 31 months, 
which have barely moved over time, underscoring the importance of accounting for censoring 
when examining birth spacing.
}
The result is that most of the median birth interval durations are still at 36 months or
below, with the shortest only 29 months.

@@

% [75th percentile birth intervals]
The most substantial changes occurred in the 75th percentile birth interval durations, 
where the more the parity progression probabilities declined, the more the birth interval 
lengthened. 
For example, the probability of a fourth birth for urban women with 8--11 years 
of education and two sons and one girl has declined by almost 40 percentage points as the 
75th percentile birth interval length increased by 22 months.
Compare this with rural women of no education with a boy as their first child, for whom the 
probability of a third birth declined by fewer than six percentage points while the birth 
interval increased by only slightly more than two months.

%0.637 0.241

@@

% [Birth intervals with sex selection]
\subsection{Sex Selection and Birth Spacing}

@@

\input{../tables/fertility.tex}

@@

%
% \footnote{
% See \citet{Croft2018} for more detail.
% Because NFHS-1 and NFHS-2 did not interview all women about fertility, the number of women 
% is based on the household rosters, assuming that never-married women have had no children.
% Because of the low number of births to women aged 45 to 49 this age group was not
% used for the hazard model estimations and, therefore, is also excluded for the fertility
% rate.
% The marital total fertility rate lead to unrealistically high numbers and is, therefore,
% not presented \citep{Hoem2011,Laplante2015}.
% }

@@

% \begin{figure}[htpb]
% \centering
% \rotatebox[origin=c]{90}{\small{1972-1984}}
% \setcounter{subfigure}{-1}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \captionsetup[subfigure]{font=footnotesize,labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[No Eduction]{\includegraphics[width=\textwidth]{mortality_s2_p1_low_dummies}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{-0} 
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \captionsetup[subfigure]{font=footnotesize,labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[One to Seven Years of Education]{\includegraphics[width=\textwidth]{mortality_s2_p1_med_dummies}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \\
% \rotatebox[origin=c]{90}{\small{1985-1994}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p2_low_dummies}
%     \end{minipage}
% } 
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p2_med_dummies} 
%     \end{minipage}
% } 
% \\
% \rotatebox[origin=c]{90}{\small{1995-2004}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p3_low_dummies}
%     \end{minipage}
% } 
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p3_med_dummies} 
%     \end{minipage}
% } 
% \\
% \rotatebox[origin=c]{90}{\small{2005-2016}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p4_low_dummies}
%     \end{minipage}
% } 
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p4_med_dummies} 
%     \end{minipage}
% } 
% \caption{Predicted Probability of Second Child's Infant Mortality 
% for Women with No Education
% and Women with One to Seven Years of Education}
% \label{fig:mortality_low_med}
% \end{figure}

% \begin{figure}[htpb]
% \centering
% \rotatebox[origin=c]{90}{\small{1972-1984}}
% \setcounter{subfigure}{-1}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \captionsetup[subfigure]{font=footnotesize,labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Eight to Eleven Years of Education]{\includegraphics[width=\textwidth]{mortality_s2_p1_high_dummies}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \begin{minipage}[t]{0.39\textwidth}\vspace{-2.68cm}\centering\footnotesize{Twelve or More Years of Education}\end{minipage}
% \\
% \rotatebox[origin=c]{90}{\small{1985-1994}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p2_high_dummies}
%     \end{minipage}
% } 
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p2_highest_dummies}
%     \end{minipage}
% } 
% \\
% \rotatebox[origin=c]{90}{\small{1995-2004}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p3_high_dummies}
%     \end{minipage}
% } 
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p3_highest_dummies} 
%     \end{minipage}
% } 
% \\
% \rotatebox[origin=c]{90}{\small{2005-2016}}
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p4_high_dummies}
%     \end{minipage}
% } 
% \subfloat[Second Spell]{
%     \begin{minipage}{0.37\textwidth}
%         \includegraphics[width=\textwidth]{mortality_s2_p4_highest_dummies} 
%     \end{minipage}
% } 
% \caption{Predicted Probability of Second Child's Infant Mortality 
% for Women with Eight to Eleven of Education
% and Women with Twelve or More Years of Education}
% \label{fig:mortality_high_highest}
% \end{figure}

@@

% Background
Over the past four decades, India saw a dramatic increase in the male-to-female
sex ratio at birth as access to sex selection spread and son preference remained high. 
Simultaneously, economic growth was strong, schooling increased, and the total fertility 
rate declined to almost replacement level.

% Questions
The main question I address in this paper is how birth spacing responded to the 
significant changes in India between 1972 and 2016, particularly the spread of sex 
selection.
I also examine two related questions.
First, did changes in birth spacing make us overestimate the decline in cohort fertility?
Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?

% What has happened to spacing and why
The most substantial lengthening of birth intervals occurred among the best-educated women,
likely because of substantial use of sex selection combined with falling fertility.
Take, for example, women with 12 or more years of education who had two girls. 
As the sex ratio reached almost 80\% boys, the expected median birth interval length
increased by almost 15 months, and the 75th percentile interval length increased by 21 months.
Most of the lengthening in the long intervals came immediately after the introduction of sex 
selection in India.

@@

% [predictions of declining sex selection] 
Despite predictions that the use of sex selection would decline, there is no clear 
evidence of this. 
The most likely users of sex selection continue to show substantial male-biased sex ratios, 
although there may be some leveling off.
More concerning, the increasingly male-biased sex ratios among less-educated women suggest
that sex selection use appears to be spreading as fertility falls.

% fertility bias

@@

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months after the previous birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that repeated abortions are associated with higher infant mortality
for the child eventually born; boys born after long birth intervals to families with
well-educated mothers, only daughters, and high predicted sex ratios---the combination of
which suggests repeated sex-selective abortions---have no higher infant mortality than other
children.

% Implications 
The results here paint a less rosy picture of India's prospects for a continued reduction 
in population growth than generally accepted. 
With predicted cohort fertility still substantially higher than the period fertility rate, 
India's total fertility rate will likely stabilize or even increase as birth intervals slow 
their lengthening. 
The more successful the attempts at combatting sex selection are, the more likely an 
increase in the total fertility rate will be. 
Furthermore, the rapid decline in infant mortality risk, combined with likely future 
declines as the proportion of very short birth intervals falls, may also slow the 
reduction in population growth.

% Future research and policy

@@

% The higher return to male offspring's human capital combined with access to sex
% selection may also explain another puzzle, why is the labor force participation so
% low when fertility is low and education has increased.
% As shown here, part of the answer may be that the number of children is, in fact, not as 
% low as previously thought, both because of the increased spacing and that more 
% children survive.

@@

\onehalfspacing
\bibliographystyle{aer}
\bibliography{sex_selection_spacing}

@@

\clearpage
\newpage

\appendix

% CHANGING NUMBERING OF FIGURES AND TABLES FOR APPENDIX
\renewcommand\thefigure{\thesection.\arabic{figure}}    
\renewcommand\thetable{\thesection.\arabic{table}}

\section*{Appendices for Online Publication}

These appendices are intended for online publication.
They provide the descriptive statistics, additional
estimated duration tables, and graphs for all 
education groups and spells.

\clearpage
\newpage

\section{Characteristics of Women's Work Experiences}

\setcounter{figure}{0}
\setcounter{table}{0}

Figure \ref{fig:work_by_survey} shows the percent of married women who are 
currently working at the time of the survey by age group and education level.
No other labor force participation question is consistently available across all 
four surveys. 
Because the question refers to currently working, the percentages are lower in previous
studies.

\captionsetup[figure]{skip = 10pt}

\begin{figure}[!htpb]
\centering
\rotatebox[origin=c]{90}{\small{40 Years or Older}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{currently_working_rural_40}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{currently_working_urban_40} 
    \end{minipage}
}
\\
\rotatebox[origin=c]{90}{\small{30--39 Years Old}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{currently_working_rural_30}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{currently_working_urban_30} 
    \end{minipage}
}
\\
\rotatebox[origin=c]{90}{\small{20--29 Years Old}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{currently_working_rural_20}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{currently_working_urban_20} 
    \end{minipage}
}
\caption{Percentage of married women who were working at the time of the 
survey by age group and area of residence}
\label{fig:work_by_survey}
\end{figure}

\begin{figure}[!htpb]
\centering
\rotatebox[origin=c]{90}{\small{40 Years or Older}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_cash_rural_40}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_cash_urban_40} 
    \end{minipage}
}
\\
\rotatebox[origin=c]{90}{\small{30--39 Years Old}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_cash_rural_30}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_cash_urban_30} 
    \end{minipage}
}
\\
\rotatebox[origin=c]{90}{\small{20--29 Years Old}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_cash_rural_20}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_cash_urban_20} 
    \end{minipage}
}
\caption{Percentage of women paid cash or cash and in-kind of those women who were 
working at the time of the survey by age group and area of residence}
\label{fig:work_cash_by_survey}
\end{figure}

\begin{figure}[!htpb]
\centering
\rotatebox[origin=c]{90}{\small{40 Years or Older}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_family_rural_40}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_family_urban_40} 
    \end{minipage}
}
\\
\rotatebox[origin=c]{90}{\small{30--39 Years Old}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_family_rural_30}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_family_urban_30} 
    \end{minipage}
}
\\
\rotatebox[origin=c]{90}{\small{20--29 Years Old}}
\subfloat[Rural]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_family_rural_20}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.45\textwidth}
        \includegraphics[width=\textwidth]{work_family_urban_20} 
    \end{minipage}
}
\caption{Percentage women who worked for a family member of those 
working at the time of the survey, by age group and area of residence}
\label{fig:work_family_by_survey}
\end{figure}

\clearpage
\newpage

\section{Empirical Model Details}

\setcounter{figure}{0}
\setcounter{table}{0}

% [More formal version of empirical strategy section]

>>>>

India has experienced many positive changes over the past four decades:
The economy has grown substantially,
real wages have more than doubled,
educational attainment has increased for males and females,
and the total fertility rate has fallen to 2.2 children
\citep{Bosworth2008,Dharmalingam2014,Klasen2015,
International-Institute-for-Population-Sciences-IIPS2017}.

@@

The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, past research has failed to appreciate that the use of sex-selective abortions could 
substantially increase birth spacing. 
As detailed below, after an abortion it takes 6 months or more to reach the same point in 
the next pregnancy. 
Second, other major societal changes, such as women's greater educational attainment, higher 
household income, and a low and declining female labor force participation, all 
likely influence birth spacing. 
The combination of the apparent increasing use of sex selection and societal 
changes can significantly impact birth spacing, but we know little about how much.

@@

To investigate how birth spacing has changed, I use a competing risk hazard model with 
two exit states: The birth of a girl or the birth of a boy. 
I apply the model to Hindu women's birth histories covering 1972--2016 using data from the 
four National Family and Health Surveys (NFHS).

@@

There are three main results.

@@

The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility, although the direction is ambiguous; empirically, 
higher female wages reduce fertility, whereas higher male wages increase fertility 
\citep{Hotz1997,schultz97}.
According to economic theory, the substitution and income effects' relative strengths 
determine the effect of higher wages on fertility.
The substitute effect captures that when wages increase, time becomes more expensive, and
people, therefore, work more and spend less time on non-wage earning activities, such as
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Because women spend substantially more time on childrearing than men, the substitution
effect dominates for women while the income effect dominates for men.

@@

With increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, thereby increasing the cost of having children and lowering fertility. 
With lower expected mortality, parents need fewer births to reach the desired number of
surviving children, which, in turn, allows parents to invest more in each child.

@@

Stronger son preference has traditionally been associated with shorter spacing and worse
health outcomes for daughters when there no sons in the household
\citep{Whitworth2002,Jayachandran2011}.
Although son preference increases average fertility only marginally, differential stopping
behavior means that girls tend to end up in larger families, resulting in fewer resources
per child \citep{repetto72,clark00,Basu2010,Barcellos2014}.

@@

The closest available proxies in the data for capturing the explanations linking
fertility and birth spacing are the sex composition of previous children, mothers'
education, and area of residence. 
The remainder of the section discusses how the three variables fit into the
explanations and the predicted effects.

@@

Finally, the process of "Sanskritization" implies that as lower-caste females gain access
to education, they adopt higher-caste norms such as stronger son preference and a
retraction from the formal labor market \citep{Srinivas1956}. 
The declining female labor force participation suggests that this process still 
operates \citep{Abraham2013,Chatterjee2018}.
Thus, although the expansion in female education will likely bring in groups who initially 
have different norms, we are unlikely to see substantial changes in how women with a given 
education level behave.

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated as their
household income increase the most, and because of their presumptive use of sex selection. 
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.

@@

@@

Equation \ref{eq:hazard} is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, splitting spells into smaller intervals---here equal to three months---and 
treating them as observations, I can estimate the model using a standard multinomial 
logit model.

@@

For birth interval lengths, I estimate a set of percentile durations. 
I first calculate for each woman when there is a given percentage chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80\% parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60\% $\left(100-\frac{80}{2}\right)$ 
mark on her survival curve. 
I then add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

@@

To ensure that there are enough births for the method to work, I censor spells at 96 
months (eight years) after a woman can first give birth, equivalent to 105 months 
after the birth of the prior child.
Less than 1\% of observed births occur after the cutoff.
The final sample consists of 
\input{./tables/num_women.tex}

Direct information on sex selection is not available, so I compare periods 
based on the changes in access and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--1983, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2019}.
}
Finally, although sex selection increased even after 1994, we may have passed a 
turning point in its use in the mid-2000s \citep{Das_Gupta2009,Kumar2012,Bongaarts2013,Diamond-Smith2015}.

@@

The second group of variables consists of those expected to have an approximately 
proportional effect on the hazard.
These include the mother's age when the spell begins, the household's
land ownership, and whether it belongs to a scheduled tribe or caste.
The online appendix present the descriptive statistics.

@@

The first question I address is how birth spacing responded to the significant 
changes in India.
I begin with a broad outline of how parity progression probabilities and
sex ratios have changed over time and by groups.
I then use these results to separately discuss the changes in birth spacing for 
women who do not appear to use sex selection and for those who do.

@@

@@

@@

The parity progression and the sex ratio show two broad trends.
First, 
in line with the falling total fertility rate, the likelihood of a next birth 
has decreased over time.
The likelihood of a next birth fell more rapidly, the higher the education and the 
higher the parity.
Within a given spell and period, parity progressions are lower in urban than rural areas, 
if at least one son is present, and the more educated the mother.

Second, sex ratios have become more male-dominated for women with no sons, indicating the spread 
of sex selection. 
The percentage of boys increased more quickly, the higher the education and the higher the parity. 
There are no clear trends for the other sex compositions. 
Within a given spell and period combination and in the absence of a son, sex ratios are higher, 
the more educated the mother, and the higher the parity. Sex ratios are also higher in urban than 
in rural areas.
Some women with one son also show an unnaturally high percentage of boys, although the 
failing fertility makes these estimates noisy.

@@

A remarkably high proportion of birth intervals are still very short.
For all but the most educated, 25\% or more have their second and third child 
within 24 months of the previous birth.
These intervals are substantially below the 24 months between 
\emph{pregnancies} the WHO recommends.
Even with the more substantial lengthening of birth intervals for higher parities, the 25th
percentile duration is still around 24 months for the fourth spell for women with less than
eight years of education.

Median birth interval lengths have also increased relatively little---only three to six months 
over the four decades---compared to around 3.5 months \emph{per decade} in other countries with 
declining fertility \citep{Rutstein2011,Casterline2016}.%
\footnote{
The NFHS reports show median durations of closed birth intervals of approximately 31 months, 
which have barely moved over time, underscoring the importance of accounting for censoring 
when examining birth spacing.
}
The result is that most of the median birth interval durations are still at 36 months or
below, with the shortest only 29 months.

@@

The most substantial changes occurred in the 75th percentile birth interval durations, 
where the more the parity progression probabilities declined, the more the birth interval 
lengthened. 
For example, the probability of a fourth birth for urban women with 8--11 years 
of education and two sons and one girl has declined by almost 40 percentage points as the 
75th percentile birth interval length increased by 22 months.
Compare this with rural women of no education with a boy as their first child, for whom the 
probability of a third birth declined by fewer than six percentage points while the birth 
interval increased by only slightly more than two months.

@@

\subsection{Sex Selection and Birth Spacing}

@@

\input{./tables/fertility.tex}

@@

@@

@@

Over the past four decades, India saw a dramatic increase in the male-to-female
sex ratio at birth as access to sex selection spread and son preference remained high. 
Simultaneously, economic growth was strong, schooling increased, and the total fertility 
rate declined to almost replacement level.

The main question I address in this paper is how birth spacing responded to the 
significant changes in India between 1972 and 2016, particularly the spread of sex 
selection.
I also examine two related questions.
First, did changes in birth spacing make us overestimate the decline in cohort fertility?
Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?

The most substantial lengthening of birth intervals occurred among the best-educated women,
likely because of substantial use of sex selection combined with falling fertility.
Take, for example, women with 12 or more years of education who had two girls. 
As the sex ratio reached almost 80\% boys, the expected median birth interval length
increased by almost 15 months, and the 75th percentile interval length increased by 21 months.
Most of the lengthening in the long intervals came immediately after the introduction of sex 
selection in India.

@@

Despite predictions that the use of sex selection would decline, there is no clear 
evidence of this. 
The most likely users of sex selection continue to show substantial male-biased sex ratios, 
although there may be some leveling off.
More concerning, the increasingly male-biased sex ratios among less-educated women suggest
that sex selection use appears to be spreading as fertility falls.

@@

There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months after the previous birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that repeated abortions are associated with higher infant mortality
for the child eventually born; boys born after long birth intervals to families with
well-educated mothers, only daughters, and high predicted sex ratios---the combination of
which suggests repeated sex-selective abortions---have no higher infant mortality than other
children.

The results here paint a less rosy picture of India's prospects for a continued reduction 
in population growth than generally accepted. 
With predicted cohort fertility still substantially higher than the period fertility rate, 
India's total fertility rate will likely stabilize or even increase as birth intervals slow 
their lengthening. 
The more successful the attempts at combatting sex selection are, the more likely an 
increase in the total fertility rate will be. 
Furthermore, the rapid decline in infant mortality risk, combined with likely future 
declines as the proportion of very short birth intervals falls, may also slow the 
reduction in population growth.

@@

@@

\onehalfspacing
\bibliographystyle{aer}
\bibliography{sex_selection_spacing-final}

@@
